Characteristic polynomials of p-adic matrices

### 2017-02-04 charpoly/charpoly.tex
Previous contribution: first version.

\subsection*{Previous contributions.} Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?

@@

The second ingredient we need is reciprocal polynomials.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
We define the reciprocal polynomial of order $f$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
We denote by $A^{\rec,d}$ the matrix
whose coefficients are the reciprocal polynomials of order $d$ 
of the coefficients of $A.$ 
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
\com(1-XM)^{\rec,n-1}=&\com(X-M), \\
(\chi_M \times 1)^{\rec,n}=&(1-XM) \com(1-XM).\\
\end{eqnarray*}
\end{lem}
\begin{proof}
It all comes down to the following result:
let $A \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
then $\det (A^{\rec,1})=\det(A)^{\rec,d}.$
Indeed, one can use multilinearity of the determinant on $X^d \det(A(1/X))$
to prove this result.
It directly implies the second part of the lemma.
As for the first part, since coefficients of the comatrix we are interested in
are determinants of a matrix of polynomials of degree at most 1, the result
is also clear.
\end{proof}

>>>>

\subsection*{Previous contributions.} 
%\todo{Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?}

Since the description of Kedlaya's algorithm in
\cite{kedlaya:01a}, the computation of characteristic polynomials
over $p$-adic numbers
has become a crucial point in many counting-points algorithms.
We can cite for example \cite{harvey:07a, harvey:14a} for
algorithms to count points on hyperelliptic curves.

In most of these papers, the precision analysis usually
deals with great details on how to obtain the matrices (\textit{e.g.} of action of Frobenius)  that are involved in the point-counting schemes.
However, the computation of their characteristic polynomials is often
a little bit less thoroughly studied: some
refer to fast algorithm (using divisions), while others
refer to division-free algorithms.

In \cite{caruso-roe-vaccon:15a}, the authors have begun the application
of the theory of differential precision of \cite{caruso-roe-vaccon:14a}
to stable computation of characteristic polynomials.
They have obtained a way to express the optimal precision
on the characteristic polynomials, but have not studied 
how to achieve this optimal precision.

@@

The second ingredient we need is reciprocal polynomials.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
We define the reciprocal polynomial of order $d$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
We denote by $A^{\rec,d}$ the matrix
whose coefficients are the reciprocal polynomials of order $d$ 
of the coefficients of $A.$ 
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
\com(1-XM)^{\rec,n-1}=&\com(X-M), \\
(\chi_M \times 1)^{\rec,n}=&(1-XM) \com(1-XM).\\
\end{eqnarray*}
\end{lem}
\begin{proof}
It all comes down to the following result:
let $A \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
then $\det (A^{\rec,1})=\det(A)^{\rec,d}.$
Indeed, one can use multilinearity of the determinant on $X^d \det(A(1/X))$
to prove this result.
It directly implies the second part of the lemma.
As for the first part, since coefficients of the comatrix we are interested in
are determinants of a matrix of polynomials of degree at most 1, the result
is also clear.
\end{proof}
